---
title: "Class 9 - Halloween - BIMM 143"
author: "Abzael Verduzco"
format: pdf
---

##1. Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy <- read.csv(candy_file, row.names = 1)

head(candy)
```

>Q1. How many different candy types are in this dataset?

```{r}
dim(candy)
```
There are 85 different candy types. 

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy[,2]) 
#or 
sum(candy$fruity)
```
There are 38 fruity candies. 

##2. What is your favorate candy?

```{r}
candy["Twix",]$winpercent
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
#My favorite candy is Reeses

candy["Reese's Peanut Butter cup",]$winpercent
```
84% is its win percent value. 

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```
It has a win percentage of roughly 77% percent. 

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent

```
It has a win percentage of roughly 50 percent.

```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The variable hist in a different scale than the rest of columns in the dataset, it is not numeric.

>Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
candy$chocolate
```
1 represent that chocolate is present, and 0 represent that it is not present. 


>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent, breaks=30)
```

```{r}
library(ggplot2)
```

```{r}
ggplot(candy, aes(x = winpercent)) +
  geom_histogram(binwidth=10)
```

> Q9. Is the distribution of winpercent values symmetrical?

The distribution of win values is not symmetrical. 

>Q10. Is the center of the distribution above or below 50%?

The center of distribution is below 50%. 

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
#Chocolate
choc.inds <- as.logical(candy$chocolate)
choc.win <- candy[choc.inds, 'winpercent']
choc.win
```
```{r}
#Fruity 
fruity.inds <- as.logical(candy$fruity)
fruity.win <- candy[fruity.inds, 'winpercent']
fruity.win
```

```{r}
mean(choc.win)
mean(fruity.win)
```
On average the chocolate candy is higher than the fruit candy. 

>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win,fruity.win)
```
Then it is statistical significance. 

## 3. Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```
Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble and Jawbusters are the least liked candies in this dataset. 

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing=TRUE),], n=5)
```
The top 5 all time favorite candy types out of this set are Snickers, Kit Kat, Twix, Reese's Miniatures, and Reese's Peanut Butter Cup.


```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```

```{r}
candy %>% arrange(winpercent) %>% tail(5)
```

In my opinion I like the order method more as it I can change to decreasing True or False. 

##Define some useful colors

```{r}
mycols <- rep("gray", nrow(candy))
mycols[as.logical(candy$fruity)] <- "darkgreen"
mycols[as.logical(candy$chocolate)] <- "brown"

mycols
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, rownames(candy)) + 
  geom_col(fill=mycols)
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) + 
  geom_col(fill=mycols)
```

Now, for the first time, using this plot we can answer questions like:
> Q17. What is the worst ranked chocolate candy?

Sixlets is the worst ranked chocholate candy

> Q18. What is the best ranked fruity candy?

The worst ranked fruity candy is Nik L nip

##4. Taking a look at pricepercent
```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, size=2.0, max.overlaps = 5)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's miniature as the best option regarding price percent and winpercent. 

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
expensive <- order(candy$pricepercent, decreasing=TRUE)

head(candy[expensive,c(11,12)], n=5)
```
The most expensive is the least popular, Nik L Nip


##5 Exploring the correlation structure

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Two of the variables with the strongest anti-correlation in this plot are, being chocolate and being fruity, as well as being a bar and coming in multiple amounts (pluribus). 

> Q23. Similarly, what two variables are most positively correlated?

The most positively correlated variables are being a chocolate and win percentage is usually correlated, as well as a being a choocolate and being a bar. 

##6. Principal Component Analysis

```{r}
pca<- prcomp(candy, scale=TRUE)

summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=mycols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (darkgreen), other (black)",
       caption="Data from 538")
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The variables in the correlation that were positively correlated with each other, therefore, it is makes sense, the negative direction are also correlated to each other but mostly in chocolate, for positive PC1, it is fruity candies that are usually hard and come in a bag or box of multiple candies. 
