---
title: "Class 05"
author: "Abzael Verduzco"
format: pdf
---

## Using GGPLOT

The ggplot2 package needs to be installed as it doesn't come with R "out of the box". 

We use `install.packages()` function to do this. 

```{r}
head(cars)
```

To use ggplot I need to load it up before I can all any of the functions in the package. I do this with the `library()` function. 

```{r}
library(ggplot2)
ggplot()
```

All ggplot figures have at least 3 things: 
- data (the stuff we want to plot)
- aesthetic mapping (aes values)
- geoms

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() + 
  labs(title= "Speed and Stopping Distance of Cars",
       x="Speed (MPH)", 
       y ="Stopping Distance (ft)", 
       subtitle = "Random car brand", 
       caption = "Dataset:'car'") +
  geom_smooth(method = "lm", se=FALSE) + 
  theme_bw()
```

ggplot is not the only graphing system in R there are lots of others. There is even "base R" graphics. 

```{r}
plot(cars)
```

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)

nrow(genes)
colnames(genes)
ncol(genes)
table(genes$State)
table(genes[,"State"])

round( table(genes[,"State"]) / nrow(genes) *100, 2)

```

```{r}
p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point()
p + scale_colour_manual( values=c("blue", "gray", "red") )
```
```{r}
p + scale_colour_manual( values=c("blue", "gray", "red") ) +
  labs(subtitle="Gene Expression Changes Upon Drug Treatment", 
          x= "Control (no drug)", 
          y= "Drug Treatment")
```


```{r}
library(gapminder)
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)

ggplot(gapminder_2007) + 
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) + 
  geom_point(alpha=0.5)
```
```{r}
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, color = pop) + 
  geom_point(alpha=0.8)
```

```{r}
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, size = pop) +
  geom_point(alpha=0.5) + 
  scale_size_area(max_size = 10)
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)

ggplot(gapminder_1957) + 
  aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop) +
  geom_point(alpha=0.7) + 
  scale_size_area(max_size = 10) 
```
```{r}
gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007)

ggplot(gapminder_1957) + 
  aes(x = gdpPercap, y = lifeExp, size = pop,
                 color = continent) +
  geom_point(alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```

