---
title: "Pertussis - Mini Project"
format: pdf
---

#Investigating Pertussis Cases by Year 

>>Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.


```{r}
cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L,2020L,2021L),
  No..Reported.Pertussis.Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116)
)

cdc
```

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(x = Year, y = No..Reported.Pertussis.Cases
) +
  geom_point() +
  geom_line() +
  labs(title = "Pertussis Cases by Year")

```

# 2. A tale of two vaccines (wP & aP)

>> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes(x = Year, y = No..Reported.Pertussis.Cases
) +
  geom_point() +
  geom_line() +
  labs(title = "Pertussis Cases by Year") +
  geom_vline(xintercept = 1946, color = "blue", linetype = "dashed") +
  geom_vline(xintercept = 1996, color = "red", linetype = "dashed") + 
  geom_text(aes(x = 1946, label = "wP"), y = max(cdc$No..Reported.Pertussis.Cases), vjust = -0.5, color = "blue") +
  geom_text(aes(x = 1996, label = "aP"), y = max(cdc$No..Reported.Pertussis.Cases), color = "red")

```

I notice a large decrease after the wP vaccine and a surprising rise after the aP vaccine. 

>>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

There is a rise in the amount of cases when it was almost at 0 in 1975 and onwards, I believe this might be due to vaccine hesitancy or evolution of the bacteria, becoming resistant to the vaccines. 

# 3. Exploring CMI-PB data

```{r}
# Allows us to read, write and process JSON data
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
```

```{r}
head(subject, 3)
```

>> Q.4 Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```
60 aP and 58 wP 

>> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```
There are 79 females and 39 males

>> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$biological_sex, subject$race)
```
1 Native male 
21 Asian females and 11 Asian males
2 Black females
9 Mixed females and 2 Mixed Males
1 Pacific Islander male and female
35 white females and 20 white males
11 Unknown females and 4 Unknown males

```{r}
library(lubridate)
```


```{r}
#What is today's date?
today()
```

```{r}
#How many days have passed since new year 2000?
today() - ymd("2000-01-01")
```

```{r}
#What is this in years? 
time_length( today() - ymd("2000-01-01"), "years")
```

>> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
# Use todays date to calculate age in days
subject$age <- today() - ymd(subject$year_of_birth)

subject$age
```

```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years") ) )
```

```{r}
# wP
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

>>Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

>>Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

There is a signifcant difference based on the two graphs. 

##Joining multiple tables 

```{r}
# Assuming you have dplyr loaded
library(dplyr)
library(jsonlite)

# Complete the API URLs...
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE)
titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE)
```

>>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- left_join(specimen, subject)
dim(meta)
head(meta)
```

>>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

>>How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

>>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

#4. Examine IgG Ab titer levels

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

>> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
library(ggplot2)

ggplot(igg) +
  aes(x = MFI_normalised, y = antigen) +
  geom_boxplot() +
  xlim(0, 75) +
  facet_wrap(vars(visit), nrow = 2) +
  labs(title = " Ab Titer Levels by Antigen Levels")

```

